---
title: "XGBoost Model Performance Summary"
author: "Automated Report"
date: "`r format(Sys.Date())`"
format:
  html:
    toc: true
    theme: cosmo
  pdf:
    toc: true
    documentclass: article
    number-sections: true
execute:
  echo: false
  warning: false
  message: false
params:
  rerun: false
---

```{r setup}
library(knitr)
library(readr)
library(data.table)
library(dplyr)
library(ggplot2)
library(tidyr)
library(kableExtra)
# Resolve project root and Results directory robustly
project_root <- normalizePath("..", mustWork = FALSE)
results_dir <- "Results"
if (!dir.exists(results_dir)) {
  # Try parent Results when rendering from Report/
  if (dir.exists(file.path(project_root, "Results"))) {
    results_dir <- file.path(project_root, "Results")
  }
}
# If params$rerun is TRUE, re-run the modelling script to refresh artifacts
if (isTRUE(params$rerun)) {
  message("Re-running modelling script...")
  # Ensure sourcing from project root
  if (file.exists(file.path(project_root, "Code/Modelling.R"))) {
    owd <- getwd(); setwd(project_root)
    on.exit(setwd(owd), add = TRUE)
    source("Code/Modelling.R")
  } else if (file.exists("Code/Modelling.R")) {
    source("Code/Modelling.R")
  } else {
    stop("Could not locate Code/Modelling.R for rerun.")
  }
} else {
  message("Using existing artifacts. Set params: rerun: true to regenerate.")
}
metrics_path <- file.path(results_dir, "metrics_test.csv")
imp_path <- file.path(results_dir, "feature_importance_gain.csv")
rmse_rand_path <- file.path(results_dir, "rmse_random_values.csv")
if (!file.exists(metrics_path) || !file.exists(imp_path) || !file.exists(rmse_rand_path)) {
  stop(sprintf("Missing artifacts in '%s'. Expected files: metrics_test.csv, feature_importance_gain.csv, rmse_random_values.csv", results_dir))
}
metrics <- data.table::fread(metrics_path)
imp <- data.table::fread(imp_path)
rmse_random <- data.table::fread(rmse_rand_path)
# Compute additional derived statistics
z_score <- (metrics$rmse_random_mean - metrics$rmse_test) / metrics$rmse_random_sd
approx_p <- 1 - pnorm(z_score)  # one-sided p (model better than random)
metrics$z_score <- z_score
metrics$approx_p_value <- approx_p
```

# Overview

This report summarizes the performance of the current XGBoost regression model trained on the formatted feature matrix with the outcome scaled to the range [-1, 1].

```{r echo=FALSE}
metrics_display <- metrics %>% 
  select(rmse_test, rmse_random_mean, rmse_random_sd, improvement, z_score, approx_p_value, mae_test, r2_test, spearman_test)
# Pivot to a narrow, two-column table to avoid wide PDF tables
metrics_long <- metrics_display %>% 
  pivot_longer(cols = everything(), names_to = "Metric", values_to = "Value") %>% 
  mutate(Value = signif(Value, 4))

tbl <- knitr::kable(metrics_long, caption = "Model vs. Shuffled Baseline Metrics", align = c('l','r'))
if (knitr::is_latex_output()) {
  tbl <- tbl %>% kableExtra::kable_styling(latex_options = c("hold_position")) %>%
    kableExtra::column_spec(1, width = "8cm") %>%
    kableExtra::column_spec(2, width = "5cm")
} else {
  tbl <- tbl %>% kableExtra::kable_styling(full_width = FALSE)
}
tbl
```

## Interpretation

Compared to a random shuffle of the training labels (null baseline), the model achieved:

- Test RMSE: **`r sprintf('%.4f', metrics$rmse_test)`**
- Mean shuffled RMSE: **`r sprintf('%.4f', metrics$rmse_random_mean)`** (SD **`r sprintf('%.4f', metrics$rmse_random_sd)`**)
- Absolute improvement (lower RMSE vs. shuffled mean): **`r sprintf('%.4f', metrics$improvement)`**
- Approximate z-score (improvement / SD of shuffled RMSEs): **`r sprintf('%.2f', metrics$z_score)`**
- Approximate one-sided p-value: **`r sprintf('%.3g', metrics$approx_p_value)`** (probability a random model is at least this good under a normal approximation)

The positive improvement and low RMSE relative to the shuffled baseline indicate the model captures signal beyond chance, though the modest z-score suggests the margin is limited and further optimization or feature engineering could be beneficial.

R-squared is **`r sprintf('%.3f', metrics$r2_test)`** (can be negative for models worse than the mean baseline under scaling). Spearman correlation **`r sprintf('%.3f', metrics$spearman_test)`** provides a rank-based agreement measure.

## Feature Importance (Gain)

The top features by XGBoost gain demonstrate which predictors most contributed to reducing loss across splits.

```{r echo=FALSE}
head_imp <- imp %>% slice_head(n = 15) %>% mutate(Feature = factor(Feature, levels = rev(Feature)))
plt_imp <- ggplot(head_imp, aes(x = Feature, y = Gain)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 15 Features (Gain)", x = "Feature", y = "Gain") +
  theme_minimal()
print(plt_imp)
```

A full CSV of importance values is saved at `Results/feature_importance_gain.csv`.

## Diagnostic Figures

### Predicted vs. True

```{r echo=FALSE, fig.cap='Predicted vs True (Test Set)'}
if (file.exists(file.path(results_dir, 'pred_vs_true_test.pdf'))) {
  knitr::include_graphics(file.path(results_dir, 'pred_vs_true_test.pdf'))
} else {
  cat('Figure not found (pred_vs_true_test.pdf).')
}
```

### Shuffled RMSE Distribution

```{r echo=FALSE, fig.cap='RMSE Distribution for Shuffled-Label Models'}
if (file.exists(file.path(results_dir, 'rmse_random_distribution.pdf'))) {
  knitr::include_graphics(file.path(results_dir, 'rmse_random_distribution.pdf'))
} else {
  cat('Figure not found (rmse_random_distribution.pdf).')
}
```

### Feature Importance Top 30

```{r echo=FALSE, fig.cap='Top 30 Feature Importances (Gain)'}
if (file.exists(file.path(results_dir, 'feature_importance_top30.pdf'))) {
  knitr::include_graphics(file.path(results_dir, 'feature_importance_top30.pdf'))
} else {
  cat('Figure not found (feature_importance_top30.pdf).')
}
```

## Recommendations

1. Explore narrower tuning around best hyperparameters (e.g., eta and max_depth) to attempt incremental gains.
2. Consider feature grouping or dimensionality reduction (e.g., PCA, pathway aggregation) to amplify signal.
3. Evaluate alternative gradient boosting libraries (LightGBM, CatBoost) for potential performance improvements.
4. Investigate calibration (e.g., isotonic regression) if prediction distribution needs smoothing.

## Session Info

```{r}
sessionInfo()
```
